### 2021-10-14 text/outline/outline.tex
finish related works section

Despite their promise, single-cell CRISPR screens present substantial statistical challenges. A major difficulty is that CRISPR perturbations are unobservable and assigned stochastically to cells. As a consequence, we cannot know with certainty which cells were perturbed. Instead, we must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed barcode counts -- to ``guess'' which cells were perturbed. Using these imputed perturbation assignments, we can attempt to estimate the effect of the perturbation on gene expression. The standard approach, which we call the ``thresholding method,'' is to assign perturbation identities to cells by simply thresholding the barcode counts.

@@

Our work makes two main contributions. First, we study the thresholding method from empirical and theoretical perspectives. Notably, we demonstrate on real data that the thresholding method exhibits a bias-variance tradeoff as a function of the selected threshold, and we recover this phenomenon in precise mathematical terms in an idealized Gaussian model. Second, we introduce a new method for estimation and inference in single-cell CRISPR screens that accounts for the measurement error inherent in the experiment. The method, called \textit{GLM-EIV} (generalized linear model with errors-in-variables), implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic. Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.

@@

Motivated (in part) by the count nature of single-cell data, several authors have generalized statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions. Lin et al.\ \cite{Lin2021} developed eSVD, an extension of SVD to exponential family and curved Gaussian response distributions. eSVD is better-suited than SVD to model the positive relationship between the mean and variance of a gene's expression level, which is induced by the countedness of the data \cite{Lause2021}. Along similar lines, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, an extension of PCA that directly models gene expression counts while regressing out technical factors. We see our work as a continuation of this broad effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on regression rather than dimension reduction: we extend the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.

The closest parallels to our work the statistical methodology literature are \cite{Grun2008} \cite{Ibrahim1990}.

>>>>

Despite their promise, single-cell CRISPR screens present substantial statistical challenges. A major difficulty is that CRISPR perturbations are unobservable and assigned stochastically to cells. As a consequence, one cannot know with certainty which cells were perturbed. Instead, one must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed barcode counts -- to ``guess'' which cells were perturbed. Using these imputed perturbation assignments, one can attempt to estimate the effect of the perturbation on gene expression. The standard approach, which we call the ``thresholding method,'' is to assign perturbation identities to cells by simply thresholding the barcode counts.

@@

Our work makes two main contributions. First, we study the thresholding method from empirical and theoretical perspectives. Notably, we demonstrate on real data that the thresholding method exhibits a bias-variance tradeoff as a function of the selected threshold, and we recover this phenomenon in precise mathematical terms in an Gaussian model. Second, we introduce a new method for estimation and inference in single-cell CRISPR screens that accounts for the measurement error inherent in the data generating process. The method, called \textit{GLM-EIV} (generalized linear model with errors-in-variables), implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic. Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.

@@

Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broad class of exponential family distributions. For example, Lin, Lei, and Roeder  \cite{Lin2021} developed eSVD, an extension of SVD to exponential family and curved Gaussian response distributions. Unlike SVD, eSVD captures the relationship between the mean and variance of a gene's expression level, a phenomenon induced by the countedness and sparsity of single-cell data \cite{Lause2021}. Along similar lines, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, an extension of PCA that directly models gene expression counts while regressing out technical factors. We see our work as a continuation of this broad effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on regression rather than dimension reduction: we extend the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.

% Several applied methods for linking perturbations to changes in gene expression in single-cell CRISPR screens ; however, these methods focus on hypothesis test (rather than estimation) 
% SCEPTRE \cite{Barry2020}

The closest parallels to our work in the statistical methodology literature are Gr\"{u}n \& Leisch \cite{Grun2008} and Ibrahim \cite{Ibrahim1990}. Gr\"{u}n \& Leisch consider estimation and inference in a $k$-component mixture of GLMs. While we prefer to view the GLM-EIV model as a kind of errors-in-variables model, the GLM-EIV model is equivalent to a two-component mixture of \textit{products} of GLM densities. Ibrahim derives a method for fitting GLMs in the presence of missing-at-random covariates. Our method, by contrast, involves fitting two conditionally independent GLMs in the presence of a totally latent covariate. Thus, while Ibrahim is helpful reference, our estimation and inference tasks are more complex than his.

The genomics literature has produced several applied methods for linking perturbations to changes in gene expression in single-cell CRISPR screens: SCEPTRE \cite{Barry2020}, MIMOSCA \cite{Dixit2016}, and scMAGeCK \cite{Yang2019}. These methods in general are focused on hypothesis testing rather than estimation; none, for instance, produces a confidence interval for the effect size of a perturbation on change in gene expression. Additionally, two of these methods (MIMOSCA and scMAGeCK) use linear models to model gene expressions, disregarding the countedness and sparsity of the data.
